---
title: "Lab 2 - Data wrangling"
author: "Aiden Mansfield"
date: "June 24, 1497"
format: 
  html: default # for speed
  pdf: default # for final output

---


```{r}
#| eval: true
#| message: false

library(tidyverse)
```


# Questions

## Part 1

### Question 1

```{r}

library(dplyr)

state_count <- midwest |> 
  group_by(state) |> 
  summarize(county_numbers = n()) |> 
  arrange(desc(county_numbers))
print("Number of counties in each state in descending order of number of counties")
print(state_count)

highest <- midwest |> 
  group_by(state) |> 
  summarize(county_numbers = n()) |> 
  ungroup() |> 
  slice_max(county_numbers, n = 1)
print("Highest number of counties")
print(highest)

lowest <- midwest |> 
  group_by(state) |> 
  summarize(county_numbers = n()) |> 
  ungroup() |> 
  slice_min(county_numbers, n = 1)
print("Lowest number of counties")
print(lowest)

```


\newpage

### Question 2

```{r}
midwest |>
  group_by(county) |>
  summarise(county_repeat = n_distinct(state)) |>
  filter(county_repeat == max(county_repeat)) |>
  select(county, county_repeat)
```


\newpage

### Question 3

```{r}
#a
midwest |>
    filter(popdensity > 25000) |>
    arrange(desc(popdensity)) |>
    select(county, state, popdensity, poptotal, area)
#b
midwest |>
    filter(popdensity == max(popdensity)) |>
    select(county,state, popdensity, poptotal, area)
```


\newpage

### Question 4

```{r}
midwest |>
  summarize(
      median_density = median(popdensity),
      lower_quartile = quantile(popdensity, 0.25),
      upper_quartile = quantile(popdensity, 0.75)
  )
```
The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has population density of 1156.208 people per unit area. The middle 50% of the counties have population densities between 622.4074 and 2330 people per unit area.

\newpage

### Question 5

```{r}
midwest |>
  mutate(metro = if_else(inmetro == 1, "Yes", "No")) |>
  group_by(state) |>
  summarize(portion_urban = mean(metro == "Yes"),
            portion_not_urban = mean(metro == "No"))
```


\newpage

### Question 6

```{r}
#a
midwest |>
  select(county, state, percbelowpoverty, percollege) |>
  filter(percbelowpoverty > 45, percbelowpoverty < 50, percollege < 10)

#b
midwest |>
  select(county, state, percbelowpoverty, percollege) |>
  filter(percollege > 40)

#c
midwest |>
    select(county, state, percbelowpoverty, percollege) |>
    filter(
      (percbelowpoverty > 45 & percbelowpoverty < 50 & percollege < 10) |
        (percollege > 40)
    )

#d
midwest_mutated <- midwest |> #needed for part e
  mutate(potential_outlier = if_else(
    (percbelowpoverty > 45 & percbelowpoverty < 50 & percollege < 10) |
        (percollege > 40), "Yes", "No"
  ))|>
  select(county, state, percbelowpoverty, percollege, potential_outlier) |>
  arrange(potential_outlier)

#e
ggplot(midwest_mutated, aes(x=percollege, y=percbelowpoverty, colour=potential_outlier, shape=state)) + geom_point() +
  labs(
    title= "Percentage with a college degree versus povoerty rate\nby State in the Midwest",
    x= "% college educated",
    y= "% below poverty",
    colour= "Potential Outlier",
    shape= "State"

  )
```


\newpage

### Question 7

\newpage

### Question 8

\newpage

## Part 2


### Question 9

